Mastery Learning in Practice:\\ A Descriptive Analysis of Data from the Cognitive Tutor Algebra I Effectiveness Trial

### 2017-10-16 paper.Rnw
models galore!

\subsection{Where To?}
\begin{figure}
<<transition,fig.height=5,fig.width=6,dependson=c('data','setup')>>=
#### what's the next section each student works?
secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))

@@

trans <- trans[rowSums(trans)>50,rowSums(trans)>50]
trans <- trans[rowSums(trans)>50,rowSums(trans)>50]

@@

nsec <- nrow(trans)

plot(1,1,xlim=c(0,6),ylim=c(0,nsec),cex=0)
text(rep(1,nsec),nsec:1,rownames(trans))
text(rep(5,nsec),nsec:1,rownames(trans))
for(i in 1:nsec){
    arrows(rep(2.5,nsec),rep(nsec-i+1,nsec),rep(3.5,nsec),nsec:1,lwd=5*trans[i,]/rowSums(trans)[i])#,col=rgb(0,0,0,trans[i,]/max(trans)))
}

@
\end{figure}
\section{Notes}
[[don't forget to include these]]

>>>>

\subsection{Where To?}
\begin{figure}
<<transition,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
#### what's the next section each student works?
secOrder <- data%>%filter(is.finite(status) & is.finite(timestamp))%>%group_by(field_id,section,unit,Curriculum,overall,year,Year,Yr,state,classid2,schoolid2)%>%summarize(time=max(timestamp),status=max(status))%>%arrange(time)%>%group_by(field_id,year,Year,Yr,state,classid2,schoolid2)%>%mutate(prevSec=c(NA,section[-n()]),prevStatus=c(NA,status[-n()]),prevUnit=c(NA,unit[-n()]),nextUnit=c(unit[-1],NA))

@@

trans <- trans[rowSums(trans)>50,colSums(trans)>10]
trans <- trans[rowSums(trans)>50,colSums(trans)>10]
trans <- trans[rowSums(trans)>50,colSums(trans)>10]

@@

nsec <- nrow(trans)
uuu <- levels(secOrderCP$unitName)
uuu <- uuu[uuu%in%c(colnames(trans),rownames(trans))]
nsec <- length(uuu)

par(mar=c(.1,.1,.1,.1))
plot(1,1,xlim=c(-.5,4),ylim=c(0,nsec),cex=0,xaxt='n',yaxt='n',xlab='',ylab='')
for(i in 1:nsec) if(uuu[i]%in%rownames(trans)) text(1,nsec-i+1,uuu[i],adj=c(1,NA))
for(i in 1:nsec) if(uuu[i]%in%colnames(trans)) text(2,nsec-i+1,uuu[i],adj=c(0,NA))

for(rr in rownames(trans))
 for(cc in colnames(trans))
     if(trans[rr,cc]>20){
      arrows(1.1,nsec-which(uuu==rr)+1,1.9,nsec-which(uuu==cc)+1,lwd=5*trans[rr,cc]/sum(trans[rr,]),
       col=rgb(0,0,0,alpha=ifelse(trans[rr,cc]>100,1,trans[rr,cc]/100)))
}

\begin{figure}
<<vcs,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
load('vcMods.RData')
library(lme4)

vcFun <- function(nm){
    mod <- vcMods[[nm]]
    out <- unlist(summary(mod)$varcor)
    out <- data.frame(sig2=out,comp=names(out),stringsAsFactors=FALSE)
    out <- rbind(out,data.frame(sig2=pi^2/3,comp='resid'))
    out$state <- strsplit(nm,'_')[[1]][1]
    out$year <- strsplit(nm,'_')[[1]][2]
    out$sig2 <- out$sig2/sum(out$sig2)
    out
}

vcDat <- do.call('rbind',lapply(names(vcMods),vcFun))

yr1 <- data.frame(sig2=c(unlist(summary(vcModYr[[1]])$varcor),pi^2/3),
                          comp=c(names(summary(vcModYr[[1]])$varcor),'resid'),
                          state='Overall',
                  year='1')
yr1$sig2 <- yr1$sig2/sum(yr1$sig2)

yr2 <- data.frame(sig2=c(unlist(summary(vcModYr[[2]])$varcor),pi^2/3),
                          comp=c(names(summary(vcModYr[[2]])$varcor),'resid'),
                          state='Overall',
                  year='2')
yr2$sig2 <- yr2$sig2/sum(yr2$sig2)

vcDat <- rbind(vcDat,yr1,yr2)

vcDat$comp <- factor(vcDat$comp)
levels(vcDat$comp)=list(State='state',School='schoolid2',Class='classid2',Student='field_id',Unit='unit',Residual='resid')

vcDat$state <- factor(vcDat$state,levels=c('TX','KY','MI','Overall'))

ggplot(vcDat,aes(year,sig2,fill=comp))+geom_col()+facet_grid(~state)+scale_fill_manual(values=rev(c('white','grey','#e41a1c','#377eb8','#4daf4a','#984ea3')))+labs(x='',y='% Variance Explained',fill='',title='Model: Multilevel Logistic Unconditional')

\begin{figure}
<<xirtCP,fig.height=7,fig.width=6.5,dependson=c('data','setup')>>=
load('cpXirtMods.RData')

xirtDat <- data%>%filter(!is.na(status) & !is.na(xirt))%>%
    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state,unit,section)%>%
    summarize(status=max(status,na.rm=TRUE),nprob=n())%>%
    group_by(field_id,classid2,schoolid2,xirt,year,spec_gifted,spec_speced,state)%>%
    summarize(nprob=sum(nprob),nsec=n_distinct(section,unit),propMast=mean(status=='graduated'),
              propCP=mean(status=='changed placement'))%>%arrange(state)

xirtDat$fitted <- NA
for(st in names(cpModSt)) xirtDat$fitted[xirtDat$state==st] <-
                                  predict(cpModSt[[st]],xirtDat[xirtDat$state==st,],
                                          allow.new.levels=TRUE,re.form=~0,type='response')

xirtDat <- addOverallState(xirtDat)

xirtDat$fitted[xirtDat$state=='Overall'] <-
    predict(cpModSimp,xirtDat[xirtDat$state=='Overall',],
            allow.new.levels=TRUE,re.form=~0,type='response')

ggplot(subset(xirtDat,state!='NJ'),aes(xirt,propCP,size=nsec))+geom_point()+geom_line(aes(xirt,fitted),col='red')+facet_grid(year~state)+coord_cartesian(ylim=c(0,0.05))

\section{Notes}
[[don't forget to include these]]
